---
title: "R Crash Course"
author: "Jonny Saunders"
date: "10/11/2017"
output: 
  md_document:
    toc: true
    toc_depth: 2
---

# Why R?

* Purpose-built for stats & data analysis - syntax is straightforward, well-implemented, and relatively static
* R Community - huge amount of centrally hosted, well-designed software with documentation to match. (vs. the nightmarish MATLAB central)
* Free, Open Source - Important not only for the state and future of computing, but also for open, reproducible science because your code can be run by anyone even if they don't have a MATLAB site license.
* Typesetting - do your statistics in the same document as your manuscript, never worry about copy and pasting a number wrong.
* Plotting - ggplot2 makes MATLAB plotting look like crayon playtime


# Basic Syntax

R has some familiar and some unfamilar syntax. 

Assignment works using both the `<-` (typical) and `=` operators, they are slightly different but in most cases are interchangeable
```{r}
x <- 1
x
```

Get help with `?`
```{r}
?lm
```


The most basic data class, a list, is declared with the `c()` operator. See my lesson on [objects](https://uodatascience.github.io/R_Guide/0201_Objects.html) for more detail. some examples:
```{r}
# Vector
x <- c(1,2,3,4,5,6)
x

# Indexing uses []
x[2]

# Matrices (arrays w/ 2 dimensions)
y <- matrix(1:24, ncol=3, nrow=8)
y

# A vector can be made an array in a few ways
matrix(x, nrow=2, ncol=3)
dim(x) <- c(2,3)
```

Data frames are one of the main ways of storing data, some examples:

```{r}
# Pull an existing dataframe for an example
diamonds <- ggplot2::diamonds # the :: notation accesses a name from a package without loading the package

head(diamonds) # show the first six rows and the header of a dataframe

# Access named columns like this
diamonds$carat[1:5]

# Logical indexing is similar to MATLAB

diamonds[diamonds$carat >= 5,] 

# The additional comma is because the logical statement selects rows, a blank after a comma selects all the columns. For example, if we just wanted the first column we could
diamonds[diamonds$carat >= 5,1] 

# or equivalently
diamonds[diamonds$carat >= 5,]$carat
```

# Packages

Packages are installed from [https://cran.r-project.org](CRAN) by default using

```{r}
# install.packages("package_name")

# Install all the packages used in this demo
# note that tidyverse is a metapackage that contains a lot of sub-packages
# eg. plyr, which i will often refer to by name and load separately
# install.packages(c("rio", "ggplot2", "R.matlab", "tidyverse", binom))

```

Installed packaged are loaded into an R session using

```{r}
#library("package_name")

# for example, load the packages that we'll be using
library("ggplot2")
library("rio")
library("R.matlab")
library("reshape2")
library("plyr")
library("binom")


# If this is the first time you have loaded rio, you will be told you are missing some packages. install them too
#install_formats()
```

# Loading data

The `rio` package's `import` function makes importing other data formats easy, it wraps a bunch of other packages that handle the actual importing. In this case though, since the out files are mega-nested gnarly matlab structures, we have to use `R.matlab`'s more flexible `readMat` function.

We'll load some of Aldis' data, comparing spiking activity in mice injected with halorhodopsin with and without laser stimulation. Because loading the data using `readMat` takes forever, I'll just leave this here, unevaluated, and used a cached version of that data. This is shitty code, but I'm leaving it all in a for loop so make it linearly readable.

```{r, eval=FALSE}

# Set this to wherever the repo landed on your computer
laser_folder <- "/Users/jonny/GitHub/R_crash_course/data/laser"
laser_files <- dir(laser_folder, pattern="outPSTH*")

# Make a blank dataframe with our field names and data types
laser_data <- data.frame(channel=integer(),
                         cluster=integer(),
                         laser=logical(),
                         amplitude=integer(),
                         rep=integer(),
                         spiketime=double())

# loop through out files
for(f in laser_files){
  print(f)
  # Load the file
  file <- file.path(laser_folder, f)
  out <- R.matlab::readMat(file)
  out <- out$out
  
  # Get field names for numerical indexing with which()
  field_names <- names(out[,,1])
  
  # Laser off first
  las_off <- which(field_names == "M1OFF")
  off <- drop(out[las_off,1,1][[1]]) # Get the M1OFF field, dropping singleton dimensions
  
  # A series of manipulation functions that massage the data into into "long" format
  off <- plyr::adply(off, c(1,2)) # adply usually applies a function, but here we just want it to group values by the first two dimensions of the array (amplitude and trial) and concatenate
  off <- melt(off, id.vars=c("X1","X2"), na.rm=TRUE) # Now we want to melt this back down so the columns with multiple spike times are flattened to a single column
  off <- off[,c(1,2,4)] # Subset the dataframe, leaving behind the spike # column
  
  # Format dataframe for concatenation
  names(off) <- c("amplitude", "rep", "spiketime") # rename fields
  off <- off[order(off$amplitude, off$rep),] # reorder by amp then rep
  off$amplitude <- as.integer(off$amplitude) # changing types and adding other variables
  off$rep <- as.integer(off$rep)
  off$channel <- as.integer(substr(f,11,11))
  off$cluster <- as.integer(substr(f,13,13))
  off$laser <- TRUE
  
  # then laser on
  las_on <- which(field_names == "M1ON")
  on <- drop(out[las_on,1,1][[1]])
  on <- plyr::adply(on, c(1,2))
  on <- melt(on, id.vars=c("X1","X2"), na.rm=TRUE)
  on <- on[,c(1,2,4)]
  
  # Format dataframe for concatenation
  names(on) <- c("amplitude", "rep", "spiketime")
  on <- on[order(on$amplitude, on$rep),]
  on$amplitude <- as.integer(on$amplitude)
  on$rep <- as.integer(on$rep)
  on$channel <- as.integer(substr(f,11,11))
  on$cluster <- as.integer(substr(f,13,13))
  on$laser <- FALSE
  
  # Concatenate to dataframe
  laser_data <- rbind(laser_data, off, on)
}

save(laser_data,file="/Users/jonny/GitHub/R_crash_course/data/laser.RData")
```

```{r}
# Load the laser data to the global environment
load("/Users/jonny/GitHub/R_crash_course/data/laser.RData", .GlobalEnv)

# Now our data is in the 'long' format
head(laser_data)
```

# Manipulating Data

Once we have our data in the long format, manipulating it is trivial

```{r}
# we use ddply to summarize our data by cell (channel, cluster), amplitude, repetition, and laser on/off
# first make a cell column that just groups channel and cluster into one column
laser_data$cell <- paste(laser_data$channel,laser_data$cluster)


rep_spikes <- ddply(laser_data, .(cell, amplitude, laser, rep), summarize,
                    n_spikes = length(spiketime))
head(rep_spikes)

# Or collapsing across repetitions
spikes <- ddply(rep_spikes, .(cell, amplitude, laser), summarize,
                mean_spikes = mean(n_spikes),
                sd_spikes = sd(n_spikes))

head(spikes)

# Or by laser condition
spikes_xy <- ddply(rep_spikes, .(cell, amplitude), summarize,
                   mean_laser_spikes = mean(n_spikes[laser==TRUE]),
                   mean_nolaser_spikes = mean(n_spikes[laser==FALSE]))

```

Some examples from my behavior data
```{r}
load("/Users/jonny/GitHub/R_crash_course/data/gendat.RData", .GlobalEnv)

head(gendat)

gendat.mouse_type <- ddply(gendat,.(mouse,gentype2),plyr::summarize, 
                      meancx   = mean(correct),
                      meanresp = mean(response),
                      meantarg = mean(target),
                      cilo     = 
                        binom.confint(sum(correct),length(correct),conf.level=0.95,method="exact")[[5]],
                      cihi     =
                        binom.confint(sum(correct),length(correct),conf.level=0.95,method="exact")[[6]],
                      nobs     = length(correct))

head(gendat.mouse_type)

```


# Plots

More detail on plotting can be found in my [https://github.com/uodatascience/ggplot_guide](ggplot guide), but a quick plot shows us what we have

```{r}
g.spikes_xy <- ggplot(spikes_xy, aes(x=mean_laser_spikes, y=mean_nolaser_spikes, color=cell))+
  geom_abline(intercept=0, slope=1)+
  geom_point()+
  coord_fixed()+
  theme(legend.position = "none")
g.spikes_xy

g.spike_ranges <- ggplot(spikes, aes(x=laser, color=as.factor(cell)))+
  geom_pointrange(aes(y=mean_spikes, ymin=mean_spikes-sd_spikes, ymax=mean_spikes+sd_spikes), size=0.3)+
  facet_grid(.~amplitude, scales="free")+
  theme(legend.position = "none")
g.spike_ranges


```

the good ole raster

```{r}
g.raster <- ggplot(laser_data[laser_data$cell == "0 1",])+
  geom_point(aes(x=spiketime, y=interaction(rep,laser), color=laser), size=0.5)+
  facet_grid(amplitude~.)+
  theme(axis.text.y = element_blank())
g.raster

g.raster_all <- ggplot(laser_data)+
  geom_point(aes(x=spiketime, y=interaction(rep,amplitude,laser), color=laser), size=0.1, alpha=0.5)+
  facet_grid(cell~.)+
  theme(axis.text.y = element_blank())
g.raster_all

g.raster_side <- ggplot(laser_data)+
  geom_point(aes(x=spiketime, y=interaction(rep,amplitude), color=laser), size=0.1, alpha=0.5)+
  facet_grid(cell~laser)+
  scale_color_brewer(palette="Set1")+
  theme(axis.text.y = element_blank())
g.raster_side 

g.density <- ggplot(laser_data[laser_data$cell == "0 1",], aes(spiketime, ..count.., fill=as.factor(amplitude)))+
  geom_density(position="fill", color=NA, alpha=0.4)+
  geom_point(aes(x=spiketime,y=(700-as.numeric(interaction(rep,amplitude)))/700, color=as.factor(amplitude)))+
  facet_grid(.~laser)+
  scale_fill_brewer(palette="Set1")+
  scale_color_brewer(palette="Set1")+
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        plot.background = element_blank(),
        panel.background = element_rect(fill="gray"),
        panel.grid = element_blank(),
        legend.position = "none")
g.density
```


# Stats

All linear models use `lm` and `glm` (regression, ANOVA, etc.). (see `?lm` and `?glm`).

So, to ANCOVA firing rate by laser and amplitude...
(note: don't do this, since the spike data is nested within each cell, you can't model the data like this. use summarized (mean) data in that case.)

```{r}
# Since our amplitude isn't really a number, but it is ordered, we cast it as an ordered factor. lm will run linear-n_levels polynomial contrasts.
rep_spikes$amplitude <- as.ordered(rep_spikes$amplitude)

# We regress n_spikes on different combinations of predictors using the formula ~ notation
# Doing hierarchal regression like we should... 
# First level predictors
laser.lm1 <- lm(n_spikes ~ as.factor(laser), data=rep_spikes)
laser.lm2 <- lm(n_spikes ~ amplitude, data=rep_spikes)
laser.lm3 <- lm(n_spikes ~ cell, data=rep_spikes)

# Second level predictors
laser.lm4 <- lm(n_spikes ~ as.factor(laser) + amplitude + cell, data=rep_spikes) # Multiple regressors
laser.lm5 <- lm(n_spikes ~ as.factor(laser) * amplitude + cell, data=rep_spikes) # Multiple regressors with interaction
laser.lm6 <- lm(n_spikes ~ as.factor(laser) + amplitude * cell, data=rep_spikes)
laser.lm7 <- lm(n_spikes ~ as.factor(laser) * amplitude * cell, data=rep_spikes)

# Model comparison

anova(laser.lm1, laser.lm2, laser.lm3, laser.lm4, laser.lm5, laser.lm6, laser.lm7)

# summarize the significant ones
pander::pander(summary(laser.lm3))
summary(laser.lm6)



```

`glm` can be used for ... well generalized linear models - so if we wanted to do a logistic regression...
```{r}
# Subset generalization data to just the best mouse
gen12 <- gendat[gendat$mouse == "7012",]

# Regress the binomial correct/incorrect responses against the type of generalization
gen12.lm <- glm(correct ~ gentype2, data=gen12, family="binomial")
summary(gen12.lm)
```


You can use the `modelr::model_matrix` function to see the equation form of the formula used for the model.

```{r}
#modelr::model_matrix()
```



# firing rate by amplitude within cells
## ANOVAs, nested analysis, skillings-mac



